{\TheTitle}

### 2017-07-11 greedy_hamiltonian_second_revision/chapters/1.introduction.tex
new comments added

In the context of Hamiltonian and Lagrangian systems, recent work suggests modifications of POD to preserve {\edit some} geometric structures. Lall et al. \cite{Lall:2003iy} and Carlberg et al. \cite{Carlberg:2014ky} suggests that the reduced-order system should be identified by a Lagrangian function on a low-dimensional configuration space. In this way, the geometric structure of the original system is inherited by the reduced system. {\edit Model reduction for port-Hamiltonian systems can be found in the works of Beattie et al. \cite{Chaturantabut:2016he}, Polyuga et al. \cite{Polyuga:2010gj} and references therein. These works construct a reduced port-Hamiltonian system using Krylov or POD methods that inherit the passivity and stability of the original system.} For Hamiltonian systems, Peng et al. \cite{Peng:2014di}, using a symplectic transformation, constructs a reduced Hamiltonian, as an approximation to the Hamiltonian of the original system. As a result, the reduced system preserves the symplectic structure. Although these methods preserve {\edit the} geometric structure, they use a POD-like approach for constructing the reduced basis and are not well {\edit suited} for problems with a high-dimensional parameter domain.

>>>>

In the context of Hamiltonian and Lagrangian systems, recent work suggests modifications of POD to preserve {\edit some} geometric structures. Lall et al. \cite{Lall:2003iy} and Carlberg et al. \cite{Carlberg:2014ky} suggests that the reduced-order system should be identified by a Lagrangian function on a low-dimensional configuration space. In this way, the geometric structure of the original system is inherited by the reduced system. {\edit Model reduction for port-Hamiltonian systems can be found in the works of Beattie et al. \cite{Chaturantabut:2016he}, Polyuga et al. \cite{Polyuga:2010gj} and references therein. These works construct a reduced port-Hamiltonian system using Krylov or POD methods that inherit the passivity and stability of the original system.} For Hamiltonian systems, Peng et al. \cite{Peng:2014di}, using a symplectic transformation, constructs a reduced Hamiltonian, as an approximation to the Hamiltonian of the original system. As a result, the reduced system preserves the symplectic structure. Although these methods preserve {\edit the} geometric structure, they use a POD-like approach for constructing the reduced basis. {\blue If the numerical evaluation of the original model is computationally demanding, performing POD can be excessively expensive \cite{Quarteroni:2016wi}.}

### 2017-07-11 greedy_hamiltonian_second_revision/chapters/3.ham.tex
new comments added

{\edit 
\begin{definition}  \cite{de2006symplectic}
Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space and $E\subset V$ is a subspace. Then the symplectic complement of $E$ inside $V$ is defined as
\[
	E^{\perp} := \{ v\in V |\ \Omega(v,e) = 0,\ \forall e\in E \}
\]
\end{definition}
Note that $E \cap E^{\perp}$ is not empty in general. 
\begin{definition} \cite{de2006symplectic}
Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space. A subspace $E\subset V$ is called a Lagrangian subspace inside $V$ if $E = E^\perp$.
\end{definition}
\begin{theorem} \label{theorem:Hasy:1.5} \cite{abraham1978foundations}
Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space. If $E\subset V$ is a Lagrangian subspace then $dim(E)=\frac 1 2dim(V)$. Here $dim$ denotes the dimension of the subspace.
\end{theorem}
\begin{definition}
A basis of $(V,\Omega)$ is called orthosymplectic if it is both a symplectic basis and an orthogonal basis with respect to the classical scalar product.
\end{definition}
\begin{theorem} \label{theorem:Hasy:1.6}  \cite{mehl2009perturbation,da2003introduction}
Suppose $(V,\Omega)$ is a $2n$ dimensional symplectic vector space and $E\subset V$ is a Lagrangian subspace. Then there is an orthosymplectic basis for $V$.
\end{theorem}
\begin{proof}
{\blue A detailed proof can be found in \cite{mehl2009perturbation}. Since the proof contains an insight to the construction of symplectic bases, we are going to restate the proof.} Starting from a Lagrangain subspace in $E \subset V$ an orthosymplectic basis can be easily constructed. By Theorem \ref{theorem:Hasy:1.5} $E$ is $n$ dimensional. Suppose that $\{ e'_1,\dots, e'_n \}$ is a basis for $E$, using the classical Gram-Schmidt orthogonalization process we can construct an orthonormal basis $\{ e_1,\dots,e_n \}$. Define a new set of vectors $f_1 = \mathbb J_{2n}^Te_1$, $f_2 =\mathbb J_{2n}^T e_2$, $\dots$, $f_n= \mathbb J_{2n}^Te_n$. We have
\begin{equation}
	\langle f_i, f_j \rangle = e_i^T \mathbb J_{2n} {\mathbb J_{2n}}^T e_j = \delta_{ij}, \quad \langle f_i, e_j \rangle = e_i^T \mathbb J_{2n} e_j = 0, \quad i,j=1,\dots,n,
\end{equation}
where we used the fact that $\mathbb J_{2n} {\mathbb J_{2n}}^T = I_{2n}$ in the first identity and the second identity is due to the fact that the basis $\{ e_1,\dots,e_n \}$ forms a Lagrangian subspace. This shows that the set $\{ e_1,\dots,e_n \}\cup \{ f_1,\dots,f_n \}$ forms an orthonormal basis. Also, it can be easily verified that this is a symplectic basis. Thus $\{ e_1,\dots,e_n \}\cup \{ f_1,\dots,f_n \}$ is an orthosymplectic basis.
\end{proof}
\begin{theorem} \label{theorem:Hasy:1.7} \cite{Marsden:1999ck}
On a finite-dimensional symplectic vector space the relationship (\ref{eq:Hasy:1}) becomes 
\begin{equation} \label{eq:Hasy:7}
\left\{
\begin{aligned}
	&\dot {\mathbf z} = \mathbb{J}_{2n} \nabla_{\mathbf z} H(\mathbf z), \\
	& \mathbf z(0) = \mathbf z_0.
\end{aligned}
\right.
\end{equation}
or, by introducing the canonical coordinates $\mathbf z = (\mathbf q^T, \mathbf p^T)^T$,
\begin{equation} \label{eq:Hasy:8}
\left\{
\begin{aligned}
	&\dot {\mathbf q} = \nabla_{\mathbf p} H(\mathbf q,\mathbf p),\\
	&\dot {\mathbf p} = - \nabla_{\mathbf q} H(\mathbf q,\mathbf p).
\end{aligned}
\right.
\end{equation}
\end{theorem}
}

@@

\begin{equation} \label{eq:Hasy:11}
	A^T \mathbb{J}_{2n}A = \mathbb{J}_{2k}.
\end{equation}
A matrix of size $2n\times 2k$ satisfying (\ref{eq:Hasy:11}) is called a \emph{symplectic matrix}.

>>>>

{\edit 
\begin{definition}  \cite{de2006symplectic}
Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space and $E\subset V$ is a subspace. Then the symplectic complement of $E$ inside $V$ is defined as
\[
	E^{\perp} := \{ v\in V |\ \Omega(v,e) = 0,\ \forall e\in E \}
\]
\end{definition}
Note that $E \cap E^{\perp}$ is not empty in general. 
\begin{definition} \cite{de2006symplectic}
Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space. A subspace $E\subset V$ is called a Lagrangian subspace inside $V$ if $E = E^\perp$.
\end{definition}
\begin{theorem} \label{theorem:Hasy:1.5} \cite{abraham1978foundations}
Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space. If $E\subset V$ is a Lagrangian subspace then $dim(E)=\frac 1 2dim(V)$. Here $dim$ denotes the dimension of the subspace.
\end{theorem}
\begin{definition}
A basis of $(V,\Omega)$ is called orthosymplectic if it is both a symplectic basis and an orthogonal basis with respect to the classical scalar product.
\end{definition}
\begin{theorem} \label{theorem:Hasy:1.6}  \cite{mehl2009perturbation,da2003introduction}
Suppose $(V,\Omega)$ is a $2n$ dimensional symplectic vector space and $E\subset V$ is a Lagrangian subspace. Then there is an orthosymplectic basis for $V$.
\end{theorem}
\begin{proof}
{\blue We are going to summarize the proof given in \cite{mehl2009perturbation}.} Starting from a Lagrangain subspace in $E \subset V$ an orthosymplectic basis can be easily constructed. By Theorem \ref{theorem:Hasy:1.5} $E$ is $n$ dimensional. Suppose that $\{ e'_1,\dots, e'_n \}$ is a basis for $E$, using the classical Gram-Schmidt orthogonalization process we can construct an orthonormal basis $\{ e_1,\dots,e_n \}$. Define a new set of vectors $f_1 = \mathbb J_{2n}^Te_1$, $f_2 =\mathbb J_{2n}^T e_2$, $\dots$, $f_n= \mathbb J_{2n}^Te_n$. We have
\begin{equation}
	\langle f_i, f_j \rangle = e_i^T \mathbb J_{2n} {\mathbb J_{2n}}^T e_j = \delta_{ij}, \quad \langle f_i, e_j \rangle = e_i^T \mathbb J_{2n} e_j = 0, \quad i,j=1,\dots,n,
\end{equation}
where we used the fact that $\mathbb J_{2n} {\mathbb J_{2n}}^T = I_{2n}$ in the first identity and the second identity is due to the fact that the basis $\{ e_1,\dots,e_n \}$ forms a Lagrangian subspace. This shows that the set $\{ e_1,\dots,e_n \}\cup \{ f_1,\dots,f_n \}$ forms an orthonormal basis. Also, it can be easily verified that this is a symplectic basis. Thus $\{ e_1,\dots,e_n \}\cup \{ f_1,\dots,f_n \}$ is an orthosymplectic basis.
\end{proof}
\begin{theorem} \label{theorem:Hasy:1.7} \cite{Marsden:1999ck}
On a finite-dimensional symplectic vector space the relationship (\ref{eq:Hasy:1}) becomes 
\begin{equation} \label{eq:Hasy:7}
\left\{
\begin{aligned}
	&\dot {\mathbf z} = \mathbb{J}_{2n} \nabla_{\mathbf z} H(\mathbf z), \\
	& \mathbf z(0) = \mathbf z_0.
\end{aligned}
\right.
\end{equation}
or, by introducing the canonical coordinates $\mathbf z = (\mathbf q^T, \mathbf p^T)^T$,
\begin{equation} \label{eq:Hasy:8}
\left\{
\begin{aligned}
	&\dot {\mathbf q} = \nabla_{\mathbf p} H(\mathbf q,\mathbf p),\\
	&\dot {\mathbf p} = - \nabla_{\mathbf q} H(\mathbf q,\mathbf p).
\end{aligned}
\right.
\end{equation}
\end{theorem}
}

@@

\begin{equation} \label{eq:Hasy:11}
	A^T \mathbb{J}_{2n}A = \mathbb{J}_{2k}.
\end{equation}
A matrix of size $2n\times 2k$ satisfying (\ref{eq:Hasy:11}) is called a \emph{symplectic matrix}. {\blue We remind the reader that a symplectic matrix is conventionally referred to a square matrix, however, here we may allow symplectic matrices to be also rectangular.}
